Short-term synaptic plasticity and threshold adaptation act in synergy to give rise to deviance detection in a spiking model of neuronal culture

### 2022-12-02 paper-1/manuscript.tex
Added remaining figure legends

Our research is based on loose approximation of a matured culture of dissociated neurons. We modeled neurons as leaky integrate-and-fire units whose membrane potential $v$ followed
\begin{equation}
    \tau_m \frac{dv}{dt} = (v_{rest}-v) + I_{syn}
\end{equation}
where
\begin{align}
    I_{syn} &= g_e(E_e-v) + g_i(E_i-v) \nonumber \\
    \tau_e \frac{dg_e}{dt} &= -g_e + \sum_{j \in \boldsymbol E} U x_j w \delta(t - \hat{t_j}) \nonumber \\
    \tau_i \frac{dg_i}{dt} &= -g_i + \sum_{j \in \boldsymbol I} w \delta(t - \hat{t_j}) \label{eqn-syn}
\end{align}
with $\boldsymbol E$ and $\boldsymbol I$ the sets of presynaptic excitatory and inhibitory neurons, respectively, $\delta(\cdot)$ the Dirac delta function, and $\hat{t}$ the time of spikes.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{Responses and threshold adaptation in a TA-only ablation. \textbf{a, b, c} Post-stimulus histograms across target trials, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev and msc sequence, respectively; right column, contrast between these two. \textbf{d} Per-neuron relationship between the contrast (dev - msc) in $V_{TA}$ at the start of trials, and the contrast in average number of spikes fired, excluding inhibitory neurons, and associated linear regression (r = -0.424, p = 3.1e-36). \textbf{e} Contribution of $V_{TA}$ and $V_{syn}$ to bins with increased firing in dev (i.e., red bins in the spike probability contrast, \textbf{a}), weighted by that same contrast. \textbf{f} Contribution of $V_{TA}$ relative to the sum of the weighted contributions in \textbf{e}. \textbf{g} Relative contribution of $V_{TA}$ across networks, showing median (solid line) and inter-quartile range (shaded area).}
    \label{fig4}
\end{figure*}

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}

@@

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation, and secondly, the more frequent presentation of B changes adaptation, reducing the average response. To estimate the effect of stimulus replacement and adaptation, respectively, we contrasted the response to B in the msc sequence with the response to other non-target stimuli in the same sequence (replacement), and with the response to B in dev (adaptation). We found that, in this particular network and stimulus pair, adaptation weakly but consistently decreased the response, whereas stimulus replacement reduced many responses more dramatically while permitting larger responses in many neurons in the general vicinity of the stimulus sites of A and B. Across networks, however, the effect of adaptation was more pronounced and more consistent, removing an average of 71 spikes per trial, whereas stimulus replacement had no effect on average (t = -0.69, p = 0.246).

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}

However, we noticed a systematic difference in the reduction of both response magnitude and $V_{TA}$ between the dev and msc conditions: The non-target response in the dev sequence was reduced more than that in the msc sequence, leading to a larger difference between the responses in these two contexts. $V_{TA}$ before A trials followed suit, leading to stronger de-adaptation and, as a consequence, a larger A trial response in dev in large parts of the network in the full model compared to TA-only. Across networks, only few cases of an increased dev response were seen; however, the response to the target stimulus (Figure \ref{fig6}e) was reduced less in dev than msc, and less in msc than std. This differential reduction was directly responsible for the larger DDI reported in Figure \ref{fig3}, but how did it arise?

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}

In Figure \ref{fig7}, we show that, in A dev and A msc trials, depression had relatively evenly distributed effects across the network. Neither sequence seemed to have a stronger overall effect, despite a slight trend towards weaker depression near the stimulus site in the dev sequence. In sharp contrast, in B dev trials, depression was very strong and very clearly localized to the stimulation site. This location dependence was closely mirrored in the B dev network response, which was largely confined to the area near the stimulation site (data not shown). To quantify this result while avoiding spurious events, we averaged the depression value $D$ over neurons with at least 1 spike per 5 trials in the given context. Across networks, we found that depression in the dev sequence was systematically stronger for B than for A (t = 7.5, p = 7.07e-12). More importantly, the level of depression in A trials changed considerably less between msc and dev sequences when compared to B trials (t = -6.7, p = 3.92e-10). Taken together, this shows that, indeed, the dominant direct effect of STD was to weaken the response to the frequently presented standard stimulus.

Finally, to complete the picture and confirm our hypothesis in full, we compared the small, but significant relief from depression in A dev trials relative to msc (t = -3.2, p = 0.001) to the difference in $V_{TA}$, applying the same neuron exclusion criteria. We found that the relief from suppression, which was shown in the TA-only analysis above to be a key ingredient to deviance detection, was driven almost exclusively by lower $V_{TA}$ (t = 11, p = 4.06e-19). More importantly, adding STD to the model induced a much larger additional difference in $V_{TA}$ compared to the TA-only condition, than in directly STD-related $D$ (t = 2.1, p = 0.0195), indicating that, indeed, STD affected the deviant response primarily through indirect effects.

>>>>

Our research is based on loose approximation of a matured culture of dissociated neurons. We modeled neurons as leaky integrate-and-fire units whose membrane potential $v$ followed
\begin{equation}
    \tau_m \frac{dv}{dt} = (v_{rest}-v) + I_{syn}
\end{equation}
where
\begin{align}
    I_{syn} &= g_e(E_e-v) + g_i(E_i-v) \nonumber \\
    \tau_e \frac{dg_e}{dt} &= -g_e + \sum_{j \in \boldsymbol E} U x_j w \delta(t - \hat{t_j}) \nonumber \\
    \tau_i \frac{dg_i}{dt} &= -g_i + \sum_{j \in \boldsymbol I} w \delta(t - \hat{t_j}) \label{eqn-syn}
\end{align}
with $\boldsymbol E$ and $\boldsymbol I$ the sets of presynaptic excitatory and inhibitory neurons, respectively, $w$ the synaptic weight, $\delta(\cdot)$ the Dirac delta function, and $\hat{t}$ the time of spikes.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{Responses and threshold adaptation in a TA-only ablation. \textbf{a, b, c} Post-stimulus histograms across target trials, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev and msc sequence, respectively; right column, contrast between these two. \textbf{d} Per-neuron relationship between the contrast (dev - msc) in $V_{TA}$ at the start of trials, and the contrast in average number of spikes fired, excluding inhibitory neurons, and associated linear regression (r = -0.424, p = 3.1e-36). \textbf{e} Contribution of $V_{TA}$ and $V_{syn}$ to bins with increased firing in dev (i.e., red bins in the spike probability contrast, \textbf{a}), weighted by that same contrast. \textbf{f} Contribution of $V_{TA}$ relative to the sum of the weighted contributions in \textbf{e}. \textbf{g} Relative contribution of $V_{TA}$ across networks, showing median (solid line) and inter-quartile range (shaded area)}
    \label{fig4}
\end{figure*}

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{\textbf{a} $V_{TA}$ at the start of A trials (mean over trials), with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence. \textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B/C/D/E in msc). \textbf{c} Left and middle: Replacement and adaptation components of the non-target response contrast (\textbf{b}, right column), where replacement is the contrast (B msc - C/D/E msc), and adaptation is the contrast (B dev - B msc). Right: Network-wide mean replacement and adaptation values across networks and target stimuli. The adaptation effect is less than zero (t = -9, p = 2.02e-15), and less than the replacement effect (t = 2.3, p = 0.0119), which is not different from zero (t = -0.69, p = 0.246). \textbf{d} Per-neuron relationship between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation, and the contrast in $V_{TA}$ at the start of target trials, excluding inhibitory neurons, and associated linear regression (r = 0.99, p = 0). \textbf{e} The same relationship as \textbf{d}, but using the medians across excitatory neurons, and plotting one point per network and target stimulus. The values are highly correlated (r = 0.956, p = 5.98e-65)}
    \label{fig5}
\end{figure*}

@@

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation, and secondly, the more frequent presentation of B changes adaptation, reducing the average response. To estimate the effect of stimulus replacement and adaptation, respectively, we contrasted the response to B in the msc sequence with the response to other non-target stimuli in the same sequence (replacement), and with the response to B in dev (adaptation). We found that, in this particular network and stimulus pair, adaptation weakly but consistently decreased the response, whereas stimulus replacement reduced many responses more dramatically while permitting larger responses in many neurons in the general vicinity of the stimulus sites of A and B. Across networks, however, the effect of adaptation was more pronounced and more consistent, removing an average of 71 spikes per trial, whereas stimulus replacement had no effect on average.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{Contrast between the ablated model with only TA (shown in Figure \ref{fig5}) and the full model with both STD and TA. \textbf{a}, \textbf{b} Data shown in Figure \ref{fig5}ab, dev and msc columns, subtracted from the corresponding data for the full model. \textbf{c} Equivalent contrast (full - ablated model) in the response to A, in columns as above. \textbf{d} Network-wide mean $V_{TA}$ before A trials across networks. Left: Ablated and full model averages. Right: Difference between these model conditions. With the introduction of STD, $V_{TA}$ decreased more in dev than in msc (t = -3.49, p = 0.000342). \textbf{e} Mean number of spikes per target trial across networks. Left: Ablated and full model averages. Right: Difference between these model conditions. With the introduction of STD, the response decreased more in the std sequence than msc (t = -1.9, p = 0.0297), and more in msc than in dev (t = -2.99, p = 0.00168)}
    \label{fig6}
\end{figure*}

However, we noticed a systematic difference in the reduction of both response magnitude and $V_{TA}$ between the dev and msc conditions: The non-target response in the dev sequence was reduced more than that in the msc sequence, leading to a larger difference between the responses in these two contexts. $V_{TA}$ before A trials followed suit, leading to stronger de-adaptation and, as a consequence, a larger A trial response in dev in large parts of the network in the full model compared to TA-only. Across networks, only few cases of an increased dev response were seen; however, the response to the target stimulus (Figure \ref{fig6}e) was reduced least in dev, and most in std. This differential reduction was directly responsible for the larger DDI reported in Figure \ref{fig3}, but how did it arise?

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{\textbf{a} Estimated average STD-derived postsynaptic depression $D$ in target (A) trials (top row) and B trials (bottom row). Note that the difference between A and B in msc (middle column) is due only to the network configuration. \textbf{b} Statistics across networks of the contrast (dev - msc) in $D$ (cf. \textbf{a}, right column), and in $V_{TA}$ at the start of A trials (cf. Figure \ref{fig6}d). In each network, averages were taken over neurons with at least 0.2 spikes per dev trial (B trials for $\Delta D^B$, A trials for $\Delta D^A$ and $\Delta V_{TA}$). Whereas the third box ($\Delta V_{TA}$ total) indicates the averages in the full model, the final box ($\Delta V_{TA}$ added) indicates the difference between the full and the ablated (TA-only) model. This additional amount of $\Delta V_{TA}$, caused indirectly by the addition of STD, was less (i.e., more negative) than the directly STD-related $\Delta D^A$ itself (t = -2.09, p = 0.0195)}
    \label{fig7}
\end{figure*}

In Figure \ref{fig7}, we show that, in A dev and A msc trials, depression had relatively evenly distributed effects across the network. Neither sequence seemed to have a stronger overall effect, despite a slight trend towards weaker depression near the stimulus site in the dev sequence. In sharp contrast, in B dev trials, depression was very strong and very clearly localized to the stimulation site. This location dependence was closely mirrored in the B dev network response, which was largely confined to the area near the stimulation site (data not shown). To quantify this result while avoiding spurious events, we averaged the depression value $D$ over neurons with at least 1 spike per 5 trials of the corresponding type in the dev sequence. Across networks, we found that depression in the dev sequence was systematically stronger for B than for A (t = 7.5, p = 7.07e-12). More importantly, the level of depression in A trials changed considerably less between msc and dev sequences when compared to B trials (t = -6.7, p = 3.92e-10). Taken together, this shows that, indeed, the dominant direct effect of STD was to weaken the response to the frequently presented standard stimulus.

Finally, to complete the picture and confirm our hypothesis in full, we compared the small, but significant relief from depression in A dev trials relative to msc (t = -3.2, p = 0.001) to the difference in $V_{TA}$, applying the same neuron exclusion criteria. We found that the relief from suppression, which was shown in the TA-only analysis above to be a key ingredient to deviance detection, was driven almost exclusively by lower $V_{TA}$ (t = 11, p = 4.06e-19). More importantly, adding STD to the model induced a much larger additional difference in $V_{TA}$ compared to the TA-only condition, than in directly STD-related $D$ (t = -2.09, p = 0.0195), indicating that, indeed, STD affected the deviant response primarily through indirect effects.
